A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-06 latex/main.tex
Updates from Overleaf

The most comprehensive analysis on recombination in SARS-CoV-2 to 
date has been performed using the combination of UShER and 
RIPPLES~\cite{Turakhia2022}. RIPPLES works by taking a
pre-existing UShER phylogeny and systematically breaking up
sequences subtended by long branches into multiple segments 
that are placed independently.
A recombination event at a node is flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This approach has much in common with the LS HMM
used in sc2ts, but with one important distinction.
While sc2ts detects and incorporates recombination
in ``real-time'' as sequences are added to the ARG in daily
batches, RIPPLES operates on a complete phylogeny,
attempting to detect recombination post-hoc among millions of nodes. 
Like any phylogenetic reconstruction method
that ignores recombination, UShER must force placement of recombinant
sequences in places that minimize the discrepancy across the entire sequence,
which may not be near any of the true parents. 
These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
may break up the artefactual long branches used to detect recombination.

@@

Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, 
and this work therefore makes several additional contributions to the field. 
First, we clarify the relationships among
recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Second, we use the ARG to study breakpoint intervals 
for recombination events, add a novel temporal dimension
and show a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage heterozygosity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we  provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
Together, these  underscore the utility of the ARG itself, and suggest
several important directions for future research.

@@

We have focused on applying sc2ts to the Viridian dataset rather 
than the larger GISAID resource~\cite{Shu2017} for two reasons.
Firstly, the Viridian dataset is of higher quality and major efforts
have been made to remove sources of error. Applying sc2ts to 
GISAID data is straightforward, but significant additional efforts 
would be required to deal with the various data quality issues.
Secondly, the Viridian data is free to download without restrictions,
enabling us to share the resulting ARG in the same manner.

@@

A set of problematic sites
for SARS-CoV-2 that are enriched for systematic sequencing errors or 
high levels of homoplasy was compiled and updated during 
the pandemic~\cite{DeMaio2020issues,Turakhia2020stability}. 
Although these sites are recommended to be masked for UShER
inference~\cite{Turakhia2020},
the list has not been updated to take into account
the systematic improvements made by the Viridian dataset, and
we therefore assembled a custom set of sites to mask. 
To do so, we first built a preliminary ARG
from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites
in terms of mutation counts in the ARG.
Only 11 of these sites
(635, 8835, 11074, 11083, 15521, 16887,
21304, 21305, 21575, 21987 and 28253)
were present in the problematic sites file
(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2}),
and most sites in this list were typical
in terms of the number of mutations.
We therefore masked our 100 mutation-rich sites
to remove sequencing errors and
improve matching performance in the LS HMM.

@@

\subsubsection*{Li and Stephens haplotype-copying model}

@@

\subsubsection*{Comparison of phylogenetic backbones}
We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
(Document \ref{sec:supplement_pango_lineages}). We then omit
omit X lineages, leaving 1329 lineages.
To ensure that each sample is a tip in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 lineages
which are singleton descendants of a recombination event in the ARG,
% NOTE: would have been simpler to say we just got rid of XBB and 
% descendants.
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an ``X''.
This results in a simplified ARG of 1282 samples,
each sample from a different Pango lineage.

@@

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the ``independent'' recombination events
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
which correspond to the Pango X lineages (if available).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(\url{https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv}).
For each of these Pango X lineages,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.

>>>>

The most comprehensive analysis on recombination in SARS-CoV-2 to 
date has been performed using the combination of UShER and 
RIPPLES~\cite{Turakhia2022}. 
% NOTE: not citing RIPPLES here as there is no real analysis
% in the paper itself, it's just focused on the tool
RIPPLES works by taking a
pre-existing UShER phylogeny and systematically breaking up
sequences subtended by long branches into multiple segments 
that are placed independently.
A recombination event at a node is flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This approach has much in common with the LS HMM
used in sc2ts, but with one important distinction.
While sc2ts detects and incorporates recombination
in ``real-time'' as sequences are added to the ARG in daily
batches, RIPPLES operates on a complete phylogeny,
attempting to detect recombination post-hoc among millions of nodes. 
Like any phylogenetic reconstruction method
that ignores recombination, UShER must force placement of recombinant
sequences in places that minimize the discrepancy across the entire sequence,
which may not be near any of the true parents. 
These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
may break up the artefactual long branches used to detect recombination.

@@

Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, 
which provides several further contributions to our understanding of the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Second, we use the ARG to study breakpoint intervals 
for recombination events, add a novel temporal dimension
and show a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we  provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
Together, these  underscore the utility of the ARG itself, and suggest
several important directions for future research.

@@

We have focused on applying sc2ts to the Viridian dataset rather 
than the larger GISAID resource~\cite{Shu2017} for two reasons.
Firstly, the Viridian dataset is of higher quality and major efforts
have been made to remove sources of error. Applying sc2ts to 
GISAID data is straightforward, but significant additional efforts 
would be required to deal with the various data quality issues.
Secondly, the Viridian data is publicly available and redistributable,
enabling us to share the resulting ARG without restrictions.

@@

A set of problematic sites
for SARS-CoV-2 that are enriched for systematic sequencing errors or 
high levels of homoplasy was compiled and updated during 
the pandemic~\cite{DeMaio2020issues,Turakhia2020stability}. 
Although these sites are recommended to be masked for UShER
inference~\cite{Turakhia2020},
the list has not been updated to take into account
the systematic improvements made by the Viridian dataset, and
we therefore assembled a custom set of sites to mask. 
To do so, we first built a preliminary ARG
from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites
in terms of mutation counts in the ARG.
Only 11 of these sites
(635, 8835, 11074, 11083, 15521, 16887,
21304, 21305, 21575, 21987 and 28253)
were present in the problematic sites file
(\url{https://github.com/W-L/ProblematicSites_SARS-CoV2}),
and most sites in this list were typical
in terms of the number of mutations.
We therefore masked our 100 most mutation-rich sites
to remove sequencing errors and
improve matching performance in the LS HMM.

@@

\subsubsection*{Li and Stephens model}

@@

\subsubsection*{Comparison of phylogenetic backbones}
We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on 1286 Pango lineages comprising those which are defined by perfect ``origination events'' 
(Document \ref{sec:supplement_pango_lineages}). but which are not acknowledged recombinant lineages (i.e. Pango X lineages and their descendants).
To ensure that each sample is a tip in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 lineages
which are singleton descendants of a putative ARG recombination event,
This results in a simplified ARG of 1282 samples,
each sample from a different Pango lineage.

@@

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the ``independent'' recombination events
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
which correspond to the Pango X lineages (if available).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
% Note from Shing: 
% This is not clearly explained in the paper at all. I went to their code to see what it means, but their code is not easy to read. 
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(\url{https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv}).
For each of these Pango X lineages,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.
